---
title: "Confidence Intervals"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
---

```{r, message = FALSE}
library(dplyr)
library(ggplot2)
theme_set(theme_bw())
library(patchwork)
library(survival)
library(mgcv)
library(pammtools)
library(batchtools)
knitr::opts_chunk$set(autodep = TRUE)
# bibliography
library(RefManageR)
BibOptions(check.entries = FALSE, hyperlink=FALSE, style = "markdown",
  max.names = 1)
bib <- ReadBib("analysis/analysis.bib", check = FALSE)
```


$$
\newcommand{\ra}{\rightarrow}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\tn}[1]{\textnormal{#1}}
\newcommand{\mbf}[1]{\mathbf{#1}}
\newcommand{\nn}{\nonumber}
\newcommand{\ub}{\underbrace}
\newcommand{\tbf}[1]{\textbf{#1}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\Prob}{\mathbb{P}}
\newcommand{\bfx}{\mathbf{x}}
\newcommand{\bfX}{\mathbf{X}}
\newcommand{\bfV}{\mathbf{V}}
\newcommand{\bfB}{\mathbf{B}}
\newcommand{\bfy}{\mathbf{y}}
\newcommand{\bff}{\mathbf{f}}
\newcommand{\bsbeta}{\boldsymbol{\beta}}
\newcommand{\bsgamma}{\boldsymbol{\gamma}}
\newcommand{\bslambda}{\boldsymbol{\lambda}}
\newcommand{\bskappa}{\boldsymbol{\kappa}}
\newcommand{\bsnu}{\boldsymbol{\nu}}
\newcommand{\bfS}{\mathbf{S}}
\newcommand{\bfz}{\mathbf{z}}
\newcommand{\bfZ}{\mathbf{Z}}
\newcommand{\drm}{\mathrm{d}}
\newcommand{\tz}{\ensuremath{t_z}}
\newcommand{\tlag}{t_{\text{lag}}}
\newcommand{\tlead}{t_{\text{lead}}}
\newcommand{\mcZ}{\mathcal{Z}}
\newcommand{\tw}{\mathcal{T}_e(j)}
\newcommand{\Tw}[1]{\mathcal{T}^{#1}}
\newcommand{\tilt}{\tilde{t}}
\newcommand{\Zi}{\ensuremath{\mathcal{Z}_i(t)}}
\newcommand{\CI}{\ensuremath{C1}}
\newcommand{\CII}{\ensuremath{C2}}
\newcommand{\CIII}{\ensuremath{C3}}
\newcommand{\gCII}{\ensuremath{g_{_{\CII}}}}
\newcommand{\gCIII}{\ensuremath{g_{_{\CIII}}}}
\newcommand{\gammaEst}{\hat{\gamma}_g^r}
\newcommand{\hatEj}{\hat{e}_{j, r}}
\newcommand{\diag}{\operatorname{diag}}
\newcommand{\rpexp}{\operatorname{rpexp}}
\newcommand{\Var}{\operatorname{Var}}$$

## Motivation
Here we consider three different ways to calculate confidence intervals for
hazard rates estimated by PAMMs and derivatives thereof, i.e., the cumulative
hazard and survival probability. The three methods compared here are

1. Delta method
2. Direct transformation
3. Simulation

## Theory
PAMMs model the (log-)hazard for which standard errors can be
obtained straight forward from the theory of GAMMs
(for details see `r Citet(bib, "Wood2017")`).
Since $\hat{\bsgamma} \sim N(\bsgamma, \bfV_{\bsgamma})$, it follows that linear
transformations of $\hat{\bsgamma}$ also follow a normal distribution.
For example, the (approximate) distribution of the log-hazard for covariate specification
$\bfx$ (row vector) is given by

$$\label{eq:linear-trafo-nv}
\bfx'\hat{\bsgamma} \sim N(\bfx'\bsgamma, \bfx' \bfV_{\bsgamma}\bfx).
$$

However, often we are also interested in quantities derived from the (log)-hazard,
most notably the cumulative hazard $\Lambda(t|\bfx)$ and the survival probability
$S(t|\bfx)$ that are non-linear transformations of $\bsgamma$. Three
approaches to derive the standard errors or confidence intervals for such
transformations are common in practice:

- Delta method
- Simulation based on the posterior distribution of the coefficients
- Direct transformation of the linear predictor

In the following, the three approaches are briefly described in more detail.
Results of a simulation study with respect to the coverage of the CI obtained by
the different approaches, are presented in section "Evaluation".

### Delta Method
Let $\hat{\bsgamma}$  the $P\times 1$ vector of coefficient estimates
and $h:\R^P \ra \R^m$. Here $m$ is the number of rows in $\bfX$.
The Delta rule states, that the transformation $h(\hat{\bsgamma})$
has normal distribution \eqref{eq:delta-rule}
\begin{equation}\label{eq:delta-rule}
h(\hat{\bsgamma}) \sim N\left(h(\bsgamma), \bfV_{h(\bsgamma)}
  :=\nabla h(\bsgamma)(\bfV_{\bsgamma}) \nabla h(\bsgamma)'\right),
\end{equation}
where $\nabla h(\bsgamma)$ is the Jacobi matrix.

Below, the variances $\bfV_{h(\bsgamma)}$ are derived for the transformations

- $h(\bsgamma) := \Lambda(t; \bfX, \bsgamma)$ and

- $h(\bsgamma) := S(t; \bfX, \bsgamma) = \exp(-\Lambda(t|\bfX, \bsgamma))$

In the following, let $t \equiv \kappa_J$ (the end point of the last interval) without loss of generality
and $\bfX$ the $J\times P$ design matrix, such that the log-hazard in
intervals $j=1,\ldots,J$ is given
by $\bs{\eta} = \bfx_1'\bsgamma,\ldots, \bfx_J'\bsgamma = \eta_1,\ldots \eta_J$.
Let further $\mbf{e}^{\mbf{v}}:=(\exp(v_1), \ldots,\exp(v_J))'$ and
$\mbf{E}^{\mbf{v}}$  the respective diagonal matrix with elements
$\mbf{E}^{\mbf{v}}_{j,j}=\exp(v_j), j=1,\ldots,J$ and
$\mbf{E}^{\mbf{v}}_{j,j'}=0\ \forall j'\neq j$.

#### Cumulative hazard

In the context of PAMMs, the cumulative hazard at time $t$ is given
by $\Lambda(t|\bfX, \bsgamma) = \sum_{j=1}^J \Delta_j \exp(\eta_j)$,
with $\Delta_j = \kappa_{j}-\kappa_{j-1}$, the length of the $j$-th interval.
In matrix notation this can be written as $\bs{\Delta} \mbf{e}^{\bs{\eta}}$,
where $\bs{\Delta}$ is the lower triangular matrix \eqref{eq:q-weights}
\begin{equation}\label{eq:q-weights}
\boldsymbol{\Delta} =
\begin{pmatrix}
\Delta_1 & \cdots & 0\\
\vdots   & \ddots & \vdots\\
\Delta_1 & \cdots & \Delta_J
\end{pmatrix}
\end{equation}

These are known constants, thus it suffices to derive the Jacobi matrix for
$\mbf{e}^{\bs{\eta}}$:

\begin{align}
\nabla \mbf{e}^{\bs{\eta}} &=
\begin{pmatrix}
\partial \frac{\exp(\bfx_1'\bsgamma)}{\partial \gamma_1} & \cdots  &\frac{\exp(\bfx_1'\bsgamma)}{\partial \gamma_{P}}\\
\vdots &\ddots & \vdots\\
\partial \frac{\exp(\bfx_J'\bsgamma)}{\partial \gamma_1} & \cdots &\frac{\exp(\bfx_J'\bsgamma)}{\partial \gamma_{P}}
\end{pmatrix}\label{eq:jacobi-hazard2}\\
 & =
 \begin{pmatrix}
 \exp(\eta_1)\cdot x_{1,1} & \cdots & \exp(\eta_1)\cdot x_{1,P}\\
 \vdots & \ddots & \vdots \\
 \exp(\eta_J)\cdot x_{J,1} & \cdots & \exp(\eta_J)\cdot x_{J,P}\\
 \end{pmatrix} \\
& = \mbf{E}^{\bs{\eta}}\bfX\label{eq:jacobi-hazard}
\end{align}

Thus the variance of
the cumulative hazard is given by

$$\begin{align}
\Var(\Lambda(t|\bfX, \bsgamma))
 & = (\bs{\Delta} \mbf{E}^{\bs{\eta}}\bfX) \bfV_{\bsgamma} (\bs{\Delta}\mbf{E}^{\bs{\eta}}\bfX)'\label{eq:V-delta-cumu}\\
 & = (\bs{\Delta} \mbf{E}^{\bs{\eta}}) (\bfX \bfV_{\bsgamma}\bfX') (\bs{\Delta}\mbf{E}^{\bs{\eta}})'\label{eq:V-delta-cumu2}
\end{align}$$

This result was also stated by `r Citet(bib, "Carstensen2005")` in a slightly
less general form.
When $t \neq \kappa_J$, $\bfX$ is a $j(t) \times P$ matrix
and $\bs{\Delta}$ is a $j(t)\times j(t)$ matrix
with elements $\Delta_1,\ldots, \Delta_{j(t)}$, where
$j(t) = j:t\in (\kappa_{j-1},\kappa_j]$ and $\Delta_{j(t)}=t-\kappa_{j(t)-1}$.

#### Survival probability

Results for the survivor function can be obtained similar to the cumulative
hazard by defining
$h(\bsgamma) := S(t|\bfX, \bsgamma)=\exp(-\Lambda(t|\bfX,\bsgamma))$.
The Jacobi matrix is then given by

$$\begin{align}
\nabla h(\bsgamma)
&= \begin{pmatrix}
 \frac{\partial e^{-\Delta_1e^{\eta_1}}}{\partial \gamma_1} & \cdots & \frac{\partial e^{-\Delta_1e^{\eta_1}}}{\partial \gamma_p}\\
 \vdots & \ddots & \vdots\\
 \frac{\partial e^{-\sum_{j=1}^J \Delta_je^{\eta_j}}}{\partial \gamma_1} & \cdots &\frac{\partial \ e^{-\sum_{j=1}^J \Delta_je^{\eta_j}}}{\partial \gamma_P}
 \end{pmatrix}\nn\\
&= \begin{pmatrix}
 e^{-\Delta_1e^{\eta_1}}\cdot(-\Delta_1e^{\eta_1})\cdot x_{1,1} & \cdots & e^{-\Delta_1e^{\eta_1}}\cdot(-\Delta_1e^{\eta_1})\cdot x_{1,P}\\
 \vdots & \ddots & \vdots\\
 e^{-\sum_{j=1}^J \Delta_je^{\eta_j}}\cdot(-\sum_{j=1}^{J}\Delta_j e^{\eta_j}\cdot x_{j,1}) & \cdots & e^{-\sum_{j=1}^J \Delta_je^{\eta_j}}\cdot(-\sum_{j=1}^{J}\Delta_j e^{\eta_j}\cdot x_{j,P})
 \end{pmatrix}\\
 & = -\mbf{E}^{-\bs{\Delta}\mbf{e}^{\bs{\eta}}}\bs{\Delta}\mbf{E}^{\bs{\eta}}\bfX\label{eq:jacobi-surv}
\end{align}
$$

The variance of the survival probability is given below:

$$
\begin{align}
\Var(S(t|\bfX, \bsgamma))
  & = (-\mbf{E}^{-\bs{\Delta}\mbf{e}^{\bs{\eta}}}\bs{\Delta}\mbf{E}^{\bs{\eta}}\bfX)\bfV_{\bsgamma} (-\mbf{E}^{-\bs{\Delta}\mbf{e}^{\bs{\eta}}}\bs{\Delta}\mbf{E}^{\bs{\eta}}\bfX)'\label{eq:V-delta-surv}\\
  & = (-\mbf{E}^{-\bs{\Delta}\mbf{e}^{\bs{\eta}}}\bs{\Delta}\mbf{E}^{\bs{\eta}})(\bfX\bfV_{\bsgamma}\bfX') (-\mbf{E}^{-\bs{\Delta}\mbf{e}^{\bs{\eta}}}\bs{\Delta}\mbf{E}^{\bs{\eta}})'\label{eq:V-delta-surv3}\\
  & = \mbf{E}^{-\bs{\Delta}\mbf{e}^{\bs{\eta}}} \Var(\Lambda(t|\bfX,\bsgamma))(\mbf{E}^{-\bs{\Delta}\mbf{e}^{\bs{\eta}}})'.\label{eq:V-delta-surv2}
\end{align}
$$

The second formulation can be used when the variance for
$\bfX\bsgamma$ is already available.
The last expression is usefull when the variance of the
cumulative hazard was obtained in a previous calculation.



### Simulation based inference
When the estimation process of a GAMM is viewed from an empirical Bayes point
of view, which, from a computational perspective, is equivalent to the REML based
approach, it can be shown
(e.g., `r Citet(bib, after = ", Ch. 7.6.1", "Fahrmeir2013")`, `r Citet(bib, after=", Ch. 4.2.4, 5.8, 6.10", "Wood2017")`),
that the posterior distribution of regression parameters $\bsgamma$ is given by

$$
\begin{equation}\label{eq:posterior-gamma}
\bsgamma|\bfy,\bsnu \sim N(\hat{\bsgamma}, \bfV_{\bsgamma})
\end{equation}
$$

with $\hat{\bsgamma}$ and $\bfV_{\bsgamma} = (\bfX'\tbf{W}\bfX + \tbf{S}_{\nu})^{-1}$
as before.
This result can be used to compute Bayesian credible intervals for any quantity
of interest that is a function of regression coefficients $\bsgamma$.
In the context of PAMMs, this approach was described by `r Citet(bib, "Argyropoulos2015")`.
For example, 95\% CIs for $\hat{S}(t|\bfx_j)$ are obtained by
drawing samples $\bsgamma_r, r=1,\ldots,R$ from the posterior,
calculating $\hat{S}_r(t,\bfx_j)$.
The lower and upper boarders of the CI is
then obtained as the $2.5\%$ and $97.5\%$ quantiles of the $R$ survival probabilities,
respectively.

### Direct transformation
One simple approach, which is often used in practice to calculating confidence intervals (CI) for monotone
transformations of the linear predictor, is to apply the
transformation to the lower and upper bound of the CI for the linear predictor.
Thus, when $\hat{\eta}_j=\bfx_j'\hat{\bsgamma}$ is the log-hazard in the $j$-th
interval with CI

$$
[\hat{l}_j=\hat{\eta}_j-\zeta_{1-\alpha/2}\hat{\sigma}_{\hat{\eta}_j},\hat{u}_j =
\hat{\eta}_j+\zeta_{1-\alpha/2}\hat{\sigma}_{\hat{\eta}_j}]
$$, with $\zeta_{1-\alpha/2}$
the $1-\alpha/2$ quantile of the normal distribution,
the CI for $h(\hat{\eta}_j)$ is obtained by $[h(\hat{l}_j); h(\hat{u}_j)]$.


## Empirical Results
### Data Simulation
We simulate data with log-hazard rate
$$
\log(\lambda(t|x)) = -3.5 + f(8,2) \cdot 6 - 0.5\cdot x_1 + \sqrt{x_2},
$$
where $f(8,2)$ is the gamma density function with respective parameters.

The below wrapper creates a data set with $n = 500$ survival times based on
above hazard. The simulation of survival times is performed using function
`pammtools::sim_pexp`.

```{r}
sim_wrapper <- function(data, job, n = 500, time_grid = seq(0, 10, by = 0.05)) {

  # create data set with covariates
  df <- tibble::tibble(x1 = runif(n, -3, 3), x2 = runif(n, 0, 6))
  # baseline hazard
  f0 <- function(t) {dgamma(t, 8, 2) * 6}

  ndf <- sim_pexp(
    formula = ~ -3.5 + f0(t) -0.5*x1 + sqrt(x2),
    data = df, cut = time_grid)

  ndf
}
```


### Estimation

The below wrapper
- transforms the simulated survival data into the
[PED format](https://adibender.github.io/pammtools/articles/data-transformation.html)
- fits the PAMM
- returns a data frame that contains the coverage for each method
- two types of splines are considered for the estimation of the baseline hazard, thinplate splines (tp)and P-Splines (ps)

<details>
  <summary>Expand to see function source code</summary>

```{r}
ci_wrapper <- function(
  data,
  job,
  instance,
  bs      = "ps",
  k       = 10,
  ci_type = "default") {

  # instance <- sim_wrapper()
  ped <- as_ped(
    data    = instance,
    formula = Surv(time, status) ~ x1 + x2,
    id      = "id")

  form <- paste0("ped_status ~ s(tend, bs='", bs, "', k=", k, ") + s(x1) + s(x2)")

  mod <- gam(
    formula = as.formula(form),
    data    = ped,
    family  = poisson(),
    offset  = offset,
    method  = "REML")

  f0 <- function(t) {dgamma(t, 8, 2) * 6}

  # create new data set
  nd <- make_newdata(ped, tend = unique(tend), x1 = c(0), x2 = c(3)) %>%
    mutate(
      true_hazard = exp(-3.5 + f0(tend) -0.5 * x1 + sqrt(x2)),
      true_cumu = cumsum(intlen * true_hazard),
      true_surv = exp(-true_cumu))
  # add hazard, cumulative hazard, survival probability with confidence intervals
  # using the 3 different methods
  nd_default <- nd  %>%
    add_hazard(mod, se_mult = qnorm(0.975), ci_type = "default") %>%
    add_cumu_hazard(mod, se_mult = qnorm(0.975), ci_type = "default") %>%
    add_surv_prob(mod, se_mult = qnorm(0.975), ci_type = "default") %>%
    mutate(
      hazard = (true_hazard >= ci_lower) & (true_hazard <= ci_upper),
      cumu = (true_cumu >= cumu_lower) & (true_cumu <= cumu_upper),
      surv =  (true_surv >= surv_lower) & (true_surv <= surv_upper)) %>%
    select(hazard, cumu, surv) %>%
    summarize_all(mean) %>%
    mutate(method = "direct")
  nd_delta <- nd %>%
    add_hazard(mod, se_mult = qnorm(0.975), ci_type = "delta", overwrite = TRUE) %>%
    add_cumu_hazard(mod, se_mult = qnorm(0.975), ci_type = "delta", overwrite = TRUE) %>%
    add_surv_prob(mod, se_mult = qnorm(0.975), ci_type = "delta", overwrite = TRUE) %>%
    mutate(
      hazard = (true_hazard >= ci_lower) & (true_hazard <= ci_upper),
      cumu = (true_cumu >= cumu_lower) & (true_cumu <= cumu_upper),
      surv =  (true_surv >= surv_lower) & (true_surv <= surv_upper)) %>%
    select(hazard, cumu, surv) %>%
    summarize_all(mean) %>%
    mutate(method = "delta")
  nd_sim <- nd %>%
    add_hazard(mod, se_mult = qnorm(0.975), ci_type = "sim", nsim = 500, overwrite = TRUE) %>%
    add_cumu_hazard(mod, se_mult = qnorm(0.975), ci_type = "sim", nsim = 500, overwrite = TRUE) %>%
    add_surv_prob(mod, se_mult = qnorm(0.975), ci_type = "sim", nsim = 500, overwrite = TRUE) %>%
    mutate(
      hazard = (true_hazard >= ci_lower) & (true_hazard <= ci_upper),
      cumu = (true_cumu >= cumu_lower) & (true_cumu <= cumu_upper),
      surv =  (true_surv >= surv_lower) & (true_surv <= surv_upper)) %>%
    select(hazard, cumu, surv) %>%
    summarize_all(mean) %>%
    mutate(method = "simulation")

    rbind(nd_default, nd_delta, nd_sim)

}
```
</details>


### Evaluation

We simulate 100 data sets and obtain respective estimates using package
**`batchtools`**:

<details>
  <summary> Expand to see **`batchtools`** code</summary>

```{r, cache = TRUE, message = FALSE, error = FALSE}
if (!checkmate::test_directory_exists("output/sim-conf-int-registry")) {
  reg <- makeExperimentRegistry(
    "output/sim-conf-int-registry",
    packages = c("mgcv", "dplyr", "tidyr", "pammtools"),
    seed     = 20052018)
  reg <- loadRegistry("output/sim-conf-int-registry", writeable = TRUE)
  # reg$cluster.functions = makeClusterFunctionsInteractive()
  addProblem(name   = "ci", fun = sim_wrapper)
  addAlgorithm(name = "ci", fun = ci_wrapper)

  algo_df <- data.frame(bs = c("tp", "ps"), stringsAsFactors = FALSE)

  addExperiments(algo.design  = list(ci = algo_df), repls = 300)

  submitJobs(findNotDone())
  # waitForJobs()

}
```

</details>

Below the RMSE and coverage are calculated for the different methods to estimate
the confidence intervals

```{r, message = FALSE}
reg     <- loadRegistry("output/sim-conf-int-registry", writeable = TRUE)
ids_res <- findExperiments(prob.name = "ci", algo.name = "ci")
pars    <- unwrap(getJobPars()) %>% as_tibble()
res     <- reduceResultsDataTable(ids=findDone(ids_res)) %>%
  as_tibble() %>%
  tidyr::unnest(cols = c(result)) %>%
  left_join(pars)
```

### Coverage table for the hazard, cumulative hazard and survival probability
```{r}
# RMSE and coverage hazard
res %>%
  group_by(bs, method) %>%
  summarize(
    "coverage hazard" = mean(hazard),
    "coverage cumulative hazard" = mean(cumu),
    "coverage survival probability" = mean(surv)) %>%
  ungroup() %>%
  mutate_if(is.numeric, ~round(., 3)) %>%
  rename("basis" = "bs") %>%
  knitr::kable()
```
