---
title: "Confidence Intervals"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: false
---

```{r, message = FALSE}
library(dplyr)
library(ggplot2)
theme_set(theme_bw())
library(patchwork)
library(survival)
library(mgcv)
library(pammtools)
library(batchtools)
knitr::opts_chunk$set(autodep = TRUE)
```

## Motivation
Here we consider three different ways to calculate confidence intervals for
hazard rates estimated by PAMMs and derivatives thereof, i.e., the cumulative
hazard and survival probability. The three methods compared here are

1. Delta method
2. Direct transformation
3. Simulation



## Data Simulation
We simulate data with log-hazard rate
$\log(\lambda(t|x)) = -3.5 + f(8,2) \cdot 6 - 0.5\cdot x_1 + \sqrt{x_2}$,
where $f(8,2)$ is the gamma density function with respective parameters.


The below wrapper creates a data set with $n = 500$ survival times based on
above hazard. The simulation of survival times is performed using function
`pammtools::sim_pexp`.

```{r}
sim_wrapper <- function(data, job, n = 500, time_grid = seq(0, 10, by = 0.05)) {

  # create data set with covariates
  df <- tibble::tibble(x1 = runif(n, -3, 3), x2 = runif(n, 0, 6))
  # baseline hazard
  f0 <- function(t) {dgamma(t, 8, 2) * 6}

  ndf <- sim_pexp(
    formula = ~ -3.5 + f0(t) -0.5*x1 + sqrt(x2),
    data = df, cut = time_grid)

  ndf
}
```


## Estimation

The below wrapper
- transforms the simulated survival data into the
[PED format](https://adibender.github.io/pammtools/articles/data-transformation.html)
- fits the PAMM
- returns a data frame that contains the coverage for each method

<details>
  <summary>Expand to see function source code</summary>

```{r}
ci_wrapper <- function(
  data,
  job,
  instance,
  bs      = "ps",
  k       = 10,
  ci_type = "default") {

  # instance <- sim_wrapper()
  ped <- as_ped(
    data    = instance,
    formula = Surv(time, status) ~ x1 + x2,
    id      = "id")

  form <- paste0("ped_status ~ s(tend, bs='", bs, "', k=", k, ") + s(x1) + s(x2)")

  mod <- gam(
    formula = as.formula(form),
    data    = ped,
    family  = poisson(),
    offset  = offset,
    method  = "REML")

  f0 <- function(t) {dgamma(t, 8, 2) * 6}

  # create new data set
  nd <- make_newdata(ped, tend = unique(tend), x1 = c(0), x2 = c(3)) %>%
    mutate(
      true_hazard = exp(-3.5 + f0(tend) -0.5 * x1 + sqrt(x2)),
      true_cumu = cumsum(intlen * true_hazard),
      true_surv = exp(-true_cumu))
  # add hazard, cumulative hazard, survival probability with confidence intervals
  # using the 3 different methods
  nd_default <- nd  %>%
    add_hazard(mod, se_mult = qnorm(0.975), ci_type = "default") %>%
    add_cumu_hazard(mod, se_mult = qnorm(0.975), ci_type = "default") %>%
    add_surv_prob(mod, se_mult = qnorm(0.975), ci_type = "default") %>%
    mutate(
      hazard = (true_hazard >= ci_lower) & (true_hazard <= ci_upper),
      cumu = (true_cumu >= cumu_lower) & (true_cumu <= cumu_upper),
      surv =  (true_surv >= surv_lower) & (true_surv <= surv_upper)) %>%
    select(hazard, cumu, surv) %>%
    summarize_all(mean) %>%
    mutate(method = "direct")
  nd_delta <- nd %>%
    add_hazard(mod, se_mult = qnorm(0.975), ci_type = "delta", overwrite = TRUE) %>%
    add_cumu_hazard(mod, se_mult = qnorm(0.975), ci_type = "delta", overwrite = TRUE) %>%
    add_surv_prob(mod, se_mult = qnorm(0.975), ci_type = "delta", overwrite = TRUE) %>%
    mutate(
      hazard = (true_hazard >= ci_lower) & (true_hazard <= ci_upper),
      cumu = (true_cumu >= cumu_lower) & (true_cumu <= cumu_upper),
      surv =  (true_surv >= surv_lower) & (true_surv <= surv_upper)) %>%
    select(hazard, cumu, surv) %>%
    summarize_all(mean) %>%
    mutate(method = "delta")
  nd_sim <- nd %>%
    add_hazard(mod, se_mult = qnorm(0.975), ci_type = "sim", nsim = 500, overwrite = TRUE) %>%
    add_cumu_hazard(mod, se_mult = qnorm(0.975), ci_type = "sim", nsim = 500, overwrite = TRUE) %>%
    add_surv_prob(mod, se_mult = qnorm(0.975), ci_type = "sim", nsim = 500, overwrite = TRUE) %>%
    mutate(
      hazard = (true_hazard >= ci_lower) & (true_hazard <= ci_upper),
      cumu = (true_cumu >= cumu_lower) & (true_cumu <= cumu_upper),
      surv =  (true_surv >= surv_lower) & (true_surv <= surv_upper)) %>%
    select(hazard, cumu, surv) %>%
    summarize_all(mean) %>%
    mutate(method = "simulation")

    rbind(nd_default, nd_delta, nd_sim)

}
```
</details>


## Evaluation

We simulate 100 data sets and obtain respective estimates using package
**`batchtools`**:

<details>
  <summary> Expand to see **`batchtools`** code</summary>

```{r, cache = TRUE, message = FALSE, error = FALSE}
if (!checkmate::test_directory_exists("output/sim-conf-int-registry")) {
  reg <- makeExperimentRegistry(
    "output/sim-conf-int-registry",
    packages = c("mgcv", "dplyr", "tidyr", "pammtools"),
    seed     = 20052018)
  reg <- loadRegistry("output/sim-conf-int-registry", writeable = TRUE)
  # reg$cluster.functions = makeClusterFunctionsInteractive()
  addProblem(name   = "ci", fun = sim_wrapper)
  addAlgorithm(name = "ci", fun = ci_wrapper)

  algo_df <- data.frame(bs = c("tp", "ps"), stringsAsFactors = FALSE)

  addExperiments(algo.design  = list(ci = algo_df), repls = 300)

  submitJobs(findNotDone())
  # waitForJobs()

}
```

</details>

Below the RMSE and coverage are calculated for the different methods to estimate
the confidence intervals

```{r, message = FALSE}
reg     <- loadRegistry("output/sim-conf-int-registry", writeable = TRUE)
ids_res <- findExperiments(prob.name = "ci", algo.name = "ci")
pars    <- unwrap(getJobPars()) %>% as_tibble()
res     <- reduceResultsDataTable(ids=findDone(ids_res)) %>%
  as_tibble() %>%
  tidyr::unnest(cols = c(result)) %>%
  left_join(pars)
```

### Coverage table for the hazard, cumulative hazard and survival probability
```{r}
# RMSE and coverage hazard
res %>%
  group_by(bs, method) %>%
  summarize(
    "coverage hazard" = mean(hazard),
    "coverage cumulative hazard" = mean(cumu),
    "coverage survival probability" = mean(surv)) %>%
  ungroup() %>%
  mutate_if(is.numeric, ~round(., 3)) %>%
  rename("basis" = "bs") %>%
  knitr::kable()
```
