---
title: "PAM Interval Endpoints vs. Midpoints"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
---
```{r setup, message = FALSE}
library(ggplot2)
theme_set(theme_bw())
library(gridExtra)
library(grid)
library(stats)
library(gtable)
library(jsonlite)
library(parallel)
library(doParallel)
library(methods)
library(Matrix)
library(Rcpp)
library(assertthat)
library(dplyr)
library(glue)
library(tibble)
library(parallel)
library(batchtools)
library(snow)
library(mgcv)
library(nlme)
library(dplyr)
library(pammtools) 
```


## Motivation
This small simulation study investigates the hazards of PAMs evaluated at
interval midpoints and how they compare to the usual approach of PAMs evaluated at interval endpoints. 

## Setup
This simulation study builds on the simulation study 
"PEM vs. PAM split point sensitivity"
and therefore uses the same data setting: 

- $n=250$ survival times are simulated from a distribution with log-hazard
$-3.5 + f(8,2)*6$, where $f(8,2)$ is the density function of the Gamma 
distribution with respective parameters.

- The baseline hazard is estimated by a PAM for interval endpoints or interval
midpoints respectively

- Three different settings are used for the interval split point definition

  1. "default": Unique event times from each simulated data set is used
  2. "fine": A fine, equidistant grid with interval lengths $0.2$
  3. "rough": A rough, equidistant grid with interval lengths $0.5$

- For each setting, 100 replications are run


Function for data simulation (using `pammtools::sim_pexp`):
```{r}
## simulation function
sim_wrapper <- function(data, job, n = 250, time_grid = seq(0, 10, by = 0.05)) {

  # create data set with covariates
  df <- tibble::tibble(x1 = runif(n, -3, 3), x2 = runif(n, 0, 6))
  # baseline hazard
  f0 <- function(t) {dgamma(t, 8, 2) * 6}
  # define function that generates nz exposures z(t_{z,1}), ..., z(t_{z,Q})

  sim_pexp(formula = ~ -3.5 + f0(t), data = df, cut = time_grid)

}
```


Function to estimate hazard from simulated data, either at interval 
endpoints or midpoints 

```{r}
## estimation function
pam_wrapper2 <- function(data, job, instance,
  cut      = NA,
  bs       = "ps",
  int_type = c("endpoint", "midpoint") ,
  max_time = 10) {

  if(is.na(cut)) {
    cut <- NULL
  } else {
    if(cut == "rough") {
      cut <- seq(0, max_time, by = 0.5)
    } else {
      if(cut == "fine") {
        cut <- seq(0, max_time, by = 0.2)
      }
    }
  }

  ped <- as_ped(data = instance, formula = Surv(time, status) ~ ., cut = cut, id="id")
  ped <- ped %>% mutate(tmid= tstart + ((tend-tstart)/2))
  
  form <- switch(int_type, 
               endpoint = ped_status ~ s(tend, bs = bs, k = 10), 
               midpoint = ped_status ~ s(tmid, bs = bs, k = 10))


  mod <- gam(formula = form, data = ped, family = poisson(), offset = offset, method = "REML")
  #summary(mod)

# Create new dataframe with hazards   
  data_seq <- data.frame(tend = seq(0.01, 10, 0.1), tmid=seq(0.01, 10, 0.1))

  data_seq <- data_seq %>%
    add_hazard(mod, type = "link", se_mult = qnorm(0.975), time_var = if(int_type == "endpoint") "tend" else "tmid") %>%
    mutate(truth = -3.5 + dgamma(if (int_type == "endpoint") tend else tmid, 8, 2) * 6)
}
```

## Setup simulation registry
Setup simulation using `batchtools`:

```{r, cache = TRUE, message = FALSE, warning=FALSE}
if(!checkmate::test_directory_exists("output/sim-pam-tmid-registry")) {
  reg <- makeExperimentRegistry("output/sim-pam-tmid-registry",
    packages = c("mgcv", "dplyr", "tidyr", "pammtools"),
    seed     = 20052018)
  reg$cluster.functions = makeClusterFunctionsSocket(ncpus = 2)
  addProblem(name   = "tend-vs-tmid", fun = sim_wrapper, seed = 20052018)
  addAlgorithm(name = "tend-vs-tmid", fun = pam_wrapper2)

  algo_df <- tidyr::crossing(
    cut = c(NA, "fine", "rough"),
    int_type = c("endpoint", "midpoint"))

  addExperiments(algo.design  = list("tend-vs-tmid" = algo_df), repls = 100)
  submitJobs() #try testJob()
  waitForJobs()
}
```


## Evaluate Simulation

```{r, message = FALSE, warning=FALSE}
reg     <- loadRegistry("output/sim-pam-tmid-registry", writeable = TRUE)
ids_pam <- findExperiments(prob.name="tend-vs-tmid", algo.name="tend-vs-tmid")
pars    <- unwrap(getJobPars()) %>% as_tibble()
res     <- reduceResultsDataTable(ids=findDone(ids_pam)) %>%
  as_tibble() %>%
  tidyr::unnest() %>%
  left_join(pars) %>%
  mutate(cut = case_when(is.na(cut) ~ "default", TRUE ~ cut))

res %>%
  mutate(
    sq_error = (truth - hazard)^2,
    covered = (truth >= ci_lower) & (truth <= ci_upper)) %>%
  group_by(job.id, int_type, cut) %>%
  summarize(
    RMSE = sqrt(mean(sq_error)),
    coverage = mean(covered)) %>%
  group_by(int_type, cut) %>%
  summarize(
    RMSE     = mean(RMSE),
    coverage = mean(coverage))

```
 .1
## Visualize Estimations
```{r, warning=FALSE}
ggplot(res, aes(x=tmid, y = hazard)) +
  geom_step(aes(group = job.id), alpha = 0.3) +
  geom_line(aes(y = truth, col = "truth"), lwd = 2) +
  facet_grid(cut ~ int_type) +
  coord_cartesian(ylim=c(-5, 0)) +
  geom_smooth(aes(col="average estimate"), method="gam", formula = y ~ s(x),
    se=FALSE) +
  scale_color_brewer("", palette = "Dark2") +
  xlab("time")

```

## Conclusion
- With an exception to the default setting, PAMs evaluated at interval endpoints rather than  midpoint seem to have a slightly higher RMSE. 
- Concerning the coverage, all of the PAMs underperformed, such that none reached
the norm coverage of 0.95. A comparison between the models reveals that the coverage for interval midpoints is slightly higher than for interval endpoints and therefore preferable. 
- Overall, PAMs at mid interval points show no substantial improvement for the coverage and might worsen the model fit. They cannot be recommended over endpoints as a general rule. 

