---
title: "PEM vs. PAM split point sensitivity"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
---

```{r setup, message = FALSE}
library(ggplot2)
theme_set(theme_bw())
library(batchtools)
```

## Motivation

This is a light-weight simulation study to investigate how sensitive the different
approaches (PEM vs. PAM) to the estimation of the baseline-hazard function are
to the placement of the interval split points.


## Setup

The setup is as follows:

- $n=250$ survival times are simulated from a distribution with log-hazard
$-3.5 + f(8,2)*6$, where $f(8,2)$ is the density function of the Gamma distribution
with respective parameters.

- The baseline hazard is estimated by a PEM and  PAM respectively

- Three different settings are used for the interval split point definition

  1. "default": Unique event times from each simulated data set is used
  2. "fine": A fine, equidistant grid with interval lengths $0.2$
  3. "rough": A rough, equidistant grid with interval lengths $0.5$

- For each setting, 20 replications are run

Function for data simulation (using `pammtools::sim_pexp`):

```{r}
## simulation function
sim_wrapper <- function(data, job, n = 250, time_grid = seq(0, 10, by = 0.05)) {

  # create data set with covariates
  df <- tibble::tibble(x1 = runif(n, -3, 3), x2 = runif(n, 0, 6))
  # baseline hazard
  f0 <- function(t) {dgamma(t, 8, 2) * 6}
  # define function that generates nz exposures z(t_{z,1}), ..., z(t_{z,Q})

  sim_pexp(formula = ~ -3.5 + f0(t), data = df, cut = time_grid)

}
```

Function to estimate hazard from simulated data, either by a PEM or PAM

```{r}
## estimation function
pam_wrapper <- function(data, job, instance,
  cut      = NA,
  bs       = "ps",
  mod_type = c("pem", "pam") ,
  max_time = 10) {

  if(is.na(cut)) {
    cut <- NULL
  } else {
    if(cut == "rough") {
      cut <- seq(0, max_time, by = 0.5)
    } else {
      if(cut == "fine") {
        cut <- seq(0, max_time, by = 0.2)
      }
    }
  }

  ped <- as_ped(data = instance, formula = Surv(time, status) ~ ., cut = cut, id="id")

  form <- "ped_status ~ s(tend) + s(x1) + s(x2)"
  if(mod_type == "pem") {
    form     <- ped_status ~ interval
    time_var <- "interval"
  } else {
    form     <- ped_status ~ s(tend, bs = bs, k = 10)
    time_var <- "tend"
  }

  mod <- gam(formula = form, data = ped, family = poisson(), offset = offset, method = "REML")
  # summary(mod)

  make_newdata(ped, tend=unique(tend)) %>%
    add_hazard(mod, type="link", se_mult = qnorm(0.975), time_var = time_var) %>%
    mutate(truth = -3.5 + dgamma(tend, 8, 2) * 6)

}
```

## Setup simulation registry
Setup simulation using `batchtools`:

```{r, cache = TRUE, message = FALSE}
if(!checkmate::test_directory_exists("output/sim-pem-vs-pam-registry")) {
  reg <- makeExperimentRegistry("output/sim-pem-vs-pam-registry",
    packages = c("mgcv", "dplyr", "tidyr", "pammtools"),
    seed     = 20052018)
  reg$cluster.functions = makeClusterFunctionsMulticore(ncpus = 2)
  addProblem(name   = "pem-vs-pam", fun = sim_wrapper)
  addAlgorithm(name = "pem-vs-pam", fun = pam_wrapper)

  algo_df <- tidyr::crossing(
    cut = c(NA, "fine", "rough"),
    mod_type = c("pem", "pam"))

  addExperiments(algo.design  = list("pem-vs-pam" = algo_df), repls = 20)
  submitJobs()
  waitForJobs()
}
```


## Evaluate Simulation

```{r, message = FALSE, warning=FALSE}
reg     <- loadRegistry("output/sim-pem-vs-pam-registry", writeable = TRUE)
ids_pam <- findExperiments(prob.name="pem-vs-pam", algo.name="pem-vs-pam")
pars    <- unwrap(getJobPars()) %>% as_tibble()
res     <- reduceResultsDataTable(ids=findDone(ids_pam)) %>%
  as_tibble() %>%
  tidyr::unnest() %>%
  left_join(pars) %>%
  mutate(cut = case_when(is.na(cut) ~ "default", TRUE ~ cut))

res %>%
  mutate(
    sq_error = (truth - hazard)^2,
    covered = (truth >= ci_lower) & (truth <= ci_upper)) %>%
  group_by(job.id, mod_type, cut) %>%
  summarize(
    RMSE = sqrt(mean(sq_error)),
    coverage = mean(covered)) %>%
  group_by(mod_type, cut) %>%
  summarize(
    RMSE     = mean(RMSE),
    coverage = mean(coverage))
```

## Visualize Estimations
```{r}
ggplot(res, aes(x=tend, y = hazard)) +
  geom_step(aes(group = job.id), alpha = 0.3) +
  geom_line(aes(y = truth, col = "truth"), lwd = 2) +
  facet_grid(cut ~ mod_type) +
  coord_cartesian(ylim=c(-5, 0)) +
  geom_smooth(aes(col="average estimate"), method="gam", formula = y ~ s(x),
    se=FALSE) +
  scale_color_brewer("", palette = "Dark2") +
  xlab("time")
```


## Conclusion

- For the PAM, the RMSE has about the same magnitude for all three split point
settings

- For the PEM, the RMSE is highly dependent on the RMSE, partly because
even for the "rough" split point setting, in some simulations some intervals
have no events and the hazard is estimated close to zero (very small log-hazard values)
and for the "default" setting, where each interval contains at least one
event, appears to overestimate the hazard on average
